### 2013-07-22 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
fixing minres -> minimal residual

In this chapter, we briefly derive the simplified $P_N$ ($SP_N$)
equations, closely following the work of Evans
\citep{evans_simpli_2013}\footnote{The $SP_N$ derivations in this
  chapter and the associated appendices are heavily based on those
  presented by Evans in \citep{evans_simpli_2013}}, in order to gain
full understanding of the underlying system and its potential behavior
in a Monte Carlo context. From the $P_N$ equations, we apply a set of
approximations to yield the $SP_N$ equations for fixed source and
criticality problems. Using the fully-formed linear operator for the
transport problem, we explore solutions to the $SP_N$ equations with
Monte Carlo Synthetic Acceleration using a light water reactor fuel
assembly criticality calculation as the driving problem. Several
difficulties arise when applying MCSA to these problems that were not
observed when investigating the simpler transport systems.

@@

\subsection{Monte Carlo Estimator Comparison}
\label{subsec:spn_estimator_comparison}
With convergence obtained for the fuel assembly problem, the advanced
preconditioners will next be studied with both the collision and
expected value estimators using the best preconditioning and
relaxation parameter combinations found by the previous
analysis. Additionally, both the Richardson-based MCSA iteration given
by Eq~(\ref{eq:mcsa}) and the MINRES-based MCSA iteration given by
Eq~(\ref{eq:mcsa_minres}) will be used. Although the expected value
estimator achieved the best iterative performance in the previous
comparison and analysis for the simple Poisson problem in
\S~\ref{sec:parameter_estimator_analysis}, it is possible that the
collision estimator may perform better from a timing perspective due
to the sparsity lost by explicit preconditioning. The deterministic
component of the expected value estimator that couples the current
state to other local states through the iteration matrix stencil
requires application of the estimator to potentially several orders of
magnitude more states at every transition event.

For this study, the fuel assembly problem with a 1-group $SP_1$
discretization was solved using MCSA with the adjoint Monte Carlo
solver using both the collision and expected value estimators and both
the Richardson and MINRES fixed point iterations. Using the results
from the previous section, a Neumann relaxation parameter of 0.7 was
used with both estimators and fixed point iterations. For the
collision estimator, a Richardson relaxation parameter of 1.1 was used
with the Richardson iteration while it was found the expected value
estimator had the best performance when a Richardson relaxation
parameter of 1.0 was used instead. In addition, ILUT preconditioning
with a drop tolerance of \sn{1}{-5} and a fill level of 5 was used
with a reduced domain approximation fill level of 100 and a threshold
of \sn{1}{-10} as in the previous analysis. Artificial absorption was
also introduced with a value of 0.2 for the MCSA calculations that
used the collision estimator as CPU timing performance was improved
without a loss in iterative performance while no absorption was used
for calculations using the expected value estimator. For each
estimator, the number of stochastic histories used to compute the MCSA
correction was varied from 25 to 50,000 with the number of MCSA
iterations required to converge to a tolerance of \sn{1}{-8} recorded
for a single eigenvalue iteration along with the CPU time required to
achieve convergence.

Figure~\ref{fig:spn_estimator_iters} gives the iteration count results
and Figure~\ref{fig:spn_estimator_time} gives the timing
results. Remarkably, although the fuel assembly problem is
significantly more complicated than the simple Poisson problem studied
in \S~\ref{sec:parameter_estimator_analysis}, the results here are
largely the same with Figure~\ref{fig:spn_estimator_iters} and
Figure~\ref{fig:estimator_nh_iters} having identical qualitative
behavior. Using the expected value estimator with both fixed point
iterations creates a relative insensitivity to the number of histories
used to compute the correction. Marginally better iterative and timing
performance was observed for the expected value estimator when using
the Richardson iteration. Better timing performance is expected as
computing the extrapolation parameter in the MINRES iteration requires
several additional parallel operations. However, we expect that
because the MINRES iteration will converge faster on its own when
compared to the Richardson iteration that better iterative performance
will be observed when it is accelerated with MCSA. The fact that this
was not observed signals that an MCSA scheme leveraging the expected
value estimator is also effectively insensitive to the fixed point
iteration used. For the collision estimator, the MINRES iteration
permits fewer stochastic histories to be used with each MCSA iteration
while still maintaining good iterative performance. With respect to
CPU time, there is a more pronounced effect due to the larger density
of states created by the explicit ILUT preconditioning. Using a
reduced domain fill level of 100 creates a situation where the
expected value estimator is significantly more expensive per history
to compute that the collision estimator due to the coupling of states.

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/estimator_iters.pdf}
  \end{center}
  \caption{\textbf{MCSA iterations required to converge the fuel
      assembly problem to a tolerance of \sn{1}{-8}.} \textit{Both the
      collision and expected value estimators were used with the
      adjoint Monte Carlo solver to compute the correction at each
      iteration. Each estimator was used with the Richardson and
      MINRES fixed point iterations within an MCSA iteration.}}
  \label{fig:spn_estimator_iters}
\end{figure}

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/estimator_time.pdf}
  \end{center}
  \caption{\textbf{Total CPU time in seconds required to converge the
      fuel assembly problem to a tolerance of \sn{1}{-8}.}
    \textit{Both the collision and expected value estimators were used
      with the adjoint Monte Carlo solver to compute the correction at
      each iteration. Each estimator was used with the Richardson and
      MINRES fixed point iterations within an MCSA iteration.}}
  \label{fig:spn_estimator_time}
\end{figure}

@@

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/estimator_convergence.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm as a function of MCSA
      iteration with 1,000 stochastic histories per iteration.}
    \textit{Both the collision and expected value estimators were used
      with the adjoint Monte Carlo solver to compute the correction at
      each iteration. Each estimator was used with the Richardson and
      MINRES fixed point iterations within an MCSA iteration.}}
  \label{fig:spn_estimator_convergence}
\end{figure}

@@

For the verification calculations, all solvers will be preconditioned
with ILUT using a drop tolerance of \sn{1}{-5} and a fill level of
5. For GMRES, no restrictions were placed on the size of the subspace
and therefore no restarts occurred. When the collision estimator was
used with MCSA, \sn{2}{4} stochastic histories were used to compute
the correction for every energy group in the problem (\sn{4}{4} and
\sn{8}{4} histories total for the 2 and 4 group calculations
respectively) corresponding to approximately 1 stochastic history per
DOF. When the expected value estimator was used, \sn{5}{2} stochastic
histories used for each energy group (\sn{1}{3} and \sn{1.5}{3}
histories total for the 2 and 4 group calculations) giving
approximately 1 stochastic history for every 5 DOFs. The reduced
domain approximation was also applied in conjunction with the ILUT
preconditioning using a fill level of 100 and a threshold value of
\sn{1}{-10} to reduce the density of states in the Monte Carlo
problem. For relaxation parameters, all MCSA computations used a
Neumann relaxation parameter of 0.7 while a Richardson relaxation of
1.1 was used with the collision estimator and 1.0 with the expected
value estimator as determined by the previous analysis of the
relaxation parameters. Table~\ref{tab:spn_solver_defs} gives
definitions for the solvers used to generate the results in the
remainder of this section. For the energy group structures,
Table~\ref{tab:spn_group_structure} gives the lower bounds of each
group in eV (an implicit upper bound of \sn{2}{6} eV is assumed for
group 0).
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{ll}\hline\hline
      \multicolumn{1}{l}{\textbf{Name}} & 
      \multicolumn{1}{l}{\textbf{Definition}} \\
      BiCGStab-ILUT & BiCGStab preconditioned with ILUT \\
      GMRES-ILUT & GMRES preconditioned with ILUT \\
      MCSA-ILUT-R-C & MCSA preconditioned with ILUT using Richardson \\ 
                    & fixed point iteration and collision estimator \\
      MCSA-ILUT-MR-C & MCSA preconditioned with ILUT using MINRES \\
                     & fixed point iteration and collision estimator \\
      MCSA-ILUT-R-EV & MCSA preconditioned with ILUT using Richardson \\
                     & fixed point iteration and expected value estimator \\
      MCSA-ILUT-MR-EV & MCSA preconditioned with ILUT using MINRES \\
                      & fixed point iteration and expected value estimator \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Solver definitions used for MCSA verification and
      performance analysis.} \textit{The ILUT preconditioner
      parameters were identical for all calculations and solvers.}}
  \label{tab:spn_solver_defs}
\end{table}
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cl}\hline\hline
      \multicolumn{1}{c}{\textbf{Number of Groups}} & 
      \multicolumn{1}{l}{\textbf{Lower Bounds (eV)}} \\
      1 & \{ \sn{1}{-5} \} \\
      2 & \{ \sn{1}{-1}, \sn{1}{-5} \} \\
      4 & \{ \sn{1}{1}, \sn{1}{0}, \sn{1}{-1}, \sn{1}{-5} \} \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Energy group lower bounds for the multigroup fuel
      assembly criticality problem in electron volts.} \textit{An
      implicit upper bound of \sn{2}{6} eV is assumed for group zero.}}
  \label{tab:spn_group_structure}
\end{table}

@@

\begin{itemize}
\item MCSA can be used to solve the asymmetric system generated by the
  multigroup $SP_N$ equations
\item MCSA has been incorporated into the Exnihilo neutronics
  production code base developed at Oak Ridge National Laboratory
\item Simple Jacobi-based preconditioning reduces the spectral radius
  of the $SP_N$ system below unity, however, this is not sufficient
  for MCSA convergence
\item As the spectral radius approaches unity, MCSA was demonstrated
  to break down with more stochastic histories required for
  convergence and more CPU time required per history
\item Light water reactor problems are difficult to solve with MCSA as
  they have large spectral radii due to the neutron scattering in the
  moderator
\item Advanced algebraic preconditioning strategies were applied to
  the $SP_N$ equations to obtain convergence with ILUT chosen for
  subsequent investigations
\item The reduced domain approximation was required to mitigate the
  memory and performance constraints of the explicit preconditioning
  strategy although both remain significant obstacles
\item MCSA relaxation parameters were developed and observed to
  enhance the time to solution for the fuel assembly criticality
  problem
\item MCSA was verified to produce the same flux distribution and
  k-eigenvalue for the fuel assembly as production Krylov methods
\item MCSA was observed to converge in fewer iterations per eigenvalue
  iteration than GMRES for the fuel assembly criticality problem and
  more then Bi-CGStab
\item The explicit preconditioning strategy creates MCSA runtimes
  $O(100)$ larger than the production Krylov methods for the fuel
  assembly criticality problem
\end{itemize}

>>>>

In this chapter, we briefly derive the simplified $P_N$ ($SP_N$)
equations, closely following the work of Evans
\citep{evans_simpli_2013}\footnote{The $SP_N$ derivations in this
  chapter and the associated appendices are heavily based on those
  presented by Evans in \citep{evans_simpli_2013}}, in order to gain
full understanding of the underlying system and its potential behavior
in a Monte Carlo context. From the $P_N$ equations, we apply a set of
approximations to yield the $SP_N$ equations for fixed source and
criticality problems. Using the fully-formed linear operator for the
transport problem, we explore solutions to the $SP_N$ equations with
Monte Carlo Synthetic Acceleration using a challenging light water
reactor fuel assembly criticality calculation as the driving
problem. Several difficulties arise when applying MCSA to these
problems that were not observed when investigating the simpler
transport systems.

@@

\subsection{Monte Carlo Estimator Comparison}
\label{subsec:spn_estimator_comparison}
With convergence obtained for the fuel assembly problem, the advanced
preconditioners will next be studied with both the collision and
expected value estimators using the best preconditioning and
relaxation parameter combinations found by the previous
analysis. Additionally, both the Richardson-based MCSA iteration given
by Eq~(\ref{eq:mcsa}) and the minimal residual-based MCSA iteration given by
Eq~(\ref{eq:mcsa_min_res}) will be used. Although the expected value
estimator achieved the best iterative performance in the previous
comparison and analysis for the simple Poisson problem in
\S~\ref{sec:parameter_estimator_analysis}, it is possible that the
collision estimator may perform better from a timing perspective due
to the sparsity lost by explicit preconditioning. The deterministic
component of the expected value estimator that couples the current
state to other local states through the iteration matrix stencil
requires application of the estimator to potentially several orders of
magnitude more states at every transition event.

For this study, the fuel assembly problem with a 1-group $SP_1$
discretization was solved using MCSA with the adjoint Monte Carlo
solver using both the collision and expected value estimators and both
the Richardson and minimal residual fixed point iterations. Using the
results from the previous section, a Neumann relaxation parameter of
0.7 was used with both estimators and fixed point iterations. For the
collision estimator, a Richardson relaxation parameter of 1.1 was used
with the Richardson iteration while it was found the expected value
estimator had the best performance when a Richardson relaxation
parameter of 1.0 was used instead. In addition, ILUT preconditioning
with a drop tolerance of \sn{1}{-5} and a fill level of 5 was used
with a reduced domain approximation fill level of 100 and a threshold
of \sn{1}{-10} as in the previous analysis. Artificial absorption was
also introduced with a value of 0.2 for the MCSA calculations that
used the collision estimator as CPU timing performance was improved
without a loss in iterative performance while no absorption was used
for calculations using the expected value estimator. For each
estimator, the number of stochastic histories used to compute the MCSA
correction was varied from 25 to 50,000 with the number of MCSA
iterations required to converge to a tolerance of \sn{1}{-8} recorded
for a single eigenvalue iteration along with the CPU time required to
achieve convergence.

Figure~\ref{fig:spn_estimator_iters} gives the iteration count results
and Figure~\ref{fig:spn_estimator_time} gives the timing
results. Remarkably, although the fuel assembly problem is
significantly more complicated than the simple Poisson problem studied
in \S~\ref{sec:parameter_estimator_analysis}, the results here are
largely the same with Figure~\ref{fig:spn_estimator_iters} and
Figure~\ref{fig:estimator_nh_iters} having identical qualitative
behavior. Using the expected value estimator with both fixed point
iterations creates a relative insensitivity to the number of histories
used to compute the correction. Marginally better iterative and timing
performance was observed for the expected value estimator when using
the Richardson iteration. Better timing performance is expected as
computing the extrapolation parameter in the minimal residual
iteration requires several additional parallel operations. However, we
expect that because the minimal residual iteration will converge
faster on its own when compared to the Richardson iteration that
better iterative performance will be observed when it is accelerated
with MCSA. The fact that this was not observed signals that an MCSA
scheme leveraging the expected value estimator is also effectively
insensitive to the fixed point iteration used. For the collision
estimator, the minimal residual iteration permits fewer stochastic
histories to be used with each MCSA iteration while still maintaining
good iterative performance. With respect to CPU time, there is a more
pronounced effect due to the larger density of states created by the
explicit ILUT preconditioning. Using a reduced domain fill level of
100 creates a situation where the expected value estimator is
significantly more expensive per history to compute that the collision
estimator due to the coupling of states.

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/estimator_iters.pdf}
  \end{center}
  \caption{\textbf{MCSA iterations required to converge the fuel
      assembly problem to a tolerance of \sn{1}{-8}.} \textit{Both the
      collision and expected value estimators were used with the
      adjoint Monte Carlo solver to compute the correction at each
      iteration. Each estimator was used with the Richardson and
      minimal residual fixed point iterations within an MCSA
      iteration.}}
  \label{fig:spn_estimator_iters}
\end{figure}

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/estimator_time.pdf}
  \end{center}
  \caption{\textbf{Total CPU time in seconds required to converge the
      fuel assembly problem to a tolerance of \sn{1}{-8}.}
    \textit{Both the collision and expected value estimators were used
      with the adjoint Monte Carlo solver to compute the correction at
      each iteration. Each estimator was used with the Richardson and
      minimal residual fixed point iterations within an MCSA
      iteration.}}
  \label{fig:spn_estimator_time}
\end{figure}

@@

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/estimator_convergence.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm as a function of MCSA
      iteration with 1,000 stochastic histories per iteration.}
    \textit{Both the collision and expected value estimators were used
      with the adjoint Monte Carlo solver to compute the correction at
      each iteration. Each estimator was used with the Richardson and
      minimal residual fixed point iterations within an MCSA
      iteration.}}
  \label{fig:spn_estimator_convergence}
\end{figure}

@@

For the verification calculations, all solvers will be preconditioned
with ILUT using a drop tolerance of \sn{1}{-5} and a fill level of
5. For GMRES, no restrictions were placed on the size of the subspace
and therefore no restarts occurred. When the collision estimator was
used with MCSA, \sn{2}{4} stochastic histories were used to compute
the correction for every energy group in the problem (\sn{4}{4} and
\sn{8}{4} histories total for the 2 and 4 group calculations
respectively) corresponding to approximately 1 stochastic history per
DOF. When the expected value estimator was used, \sn{5}{2} stochastic
histories used for each energy group (\sn{1}{3} and \sn{1.5}{3}
histories total for the 2 and 4 group calculations) giving
approximately 1 stochastic history for every 5 DOFs. The reduced
domain approximation was also applied in conjunction with the ILUT
preconditioning using a fill level of 100 and a threshold value of
\sn{1}{-10} to reduce the density of states in the Monte Carlo
problem. For relaxation parameters, all MCSA computations used a
Neumann relaxation parameter of 0.7 while a Richardson relaxation of
1.1 was used with the collision estimator and 1.0 with the expected
value estimator as determined by the previous analysis of the
relaxation parameters. Table~\ref{tab:spn_solver_defs} gives
definitions for the solvers used to generate the results in the
remainder of this section. For the energy group structures,
Table~\ref{tab:spn_group_structure} gives the lower bounds of each
group in eV (an implicit upper bound of \sn{2}{6} eV is assumed for
group 0).
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{ll}\hline\hline
      \multicolumn{1}{l}{\textbf{Name}} & 
      \multicolumn{1}{l}{\textbf{Definition}} \\
      BiCGStab-ILUT & BiCGStab preconditioned with ILUT \\
      GMRES-ILUT & GMRES preconditioned with ILUT \\
      MCSA-ILUT-R-C & MCSA preconditioned with ILUT using Richardson \\ 
                    & fixed point iteration and collision estimator \\
      MCSA-ILUT-MR-C & MCSA preconditioned with ILUT using minimal residual \\
                     & fixed point iteration and collision estimator \\
      MCSA-ILUT-R-EV & MCSA preconditioned with ILUT using Richardson \\
                     & fixed point iteration and expected value estimator \\
      MCSA-ILUT-MR-EV & MCSA preconditioned with ILUT using minimal residual \\
                      & fixed point iteration and expected value estimator \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Solver definitions used for MCSA verification and
      performance analysis.} \textit{The ILUT preconditioner
      parameters were identical for all calculations and solvers.}}
  \label{tab:spn_solver_defs}
\end{table}
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cl}\hline\hline
      \multicolumn{1}{c}{\textbf{Number of Groups}} & 
      \multicolumn{1}{l}{\textbf{Lower Bounds (eV)}} \\
      1 & \{ \sn{1}{-5} \} \\
      2 & \{ \sn{1}{-1}, \sn{1}{-5} \} \\
      4 & \{ \sn{1}{1}, \sn{1}{0}, \sn{1}{-1}, \sn{1}{-5} \} \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Energy group lower bounds for the multigroup fuel
      assembly criticality problem in electron volts.} \textit{An
      implicit upper bound of \sn{2}{6} eV is assumed for group zero.}}
  \label{tab:spn_group_structure}
\end{table}

@@

\begin{itemize}
\item MCSA can be used to solve the asymmetric system generated by the
  multigroup $SP_N$ equations
\item MCSA has been incorporated into the Exnihilo neutronics
  production code base developed at Oak Ridge National Laboratory
\item Simple Jacobi-based preconditioning reduces the spectral radius
  of the $SP_N$ system below unity, however, this is not sufficient
  for MCSA convergence
\item As the spectral radius approaches unity, MCSA was demonstrated
  to break down with more stochastic histories required for
  convergence and more CPU time required per history
\item Light water reactor problems are difficult to solve with MCSA as
  they have large spectral radii due to the neutron scattering in the
  moderator
\item Advanced algebraic preconditioning strategies were applied to
  the $SP_N$ equations to obtain convergence with ILUT chosen for
  subsequent investigations
\item The reduced domain approximation was required to mitigate the
  memory and performance constraints of the explicit preconditioning
  strategy although both remain significant obstacles
\item MCSA relaxation parameters were developed and observed to
  enhance the time to solution for the fuel assembly criticality
  problem
\item MCSA was verified to produce the same flux distribution and
  k-eigenvalue for the fuel assembly as production Krylov methods
\item MCSA was observed to converge in fewer iterations per eigenvalue
  iteration than GMRES for the fuel assembly criticality problem and
  more then Bi-CGStab
\item The explicit preconditioning strategy required to overcome the
  MCSA spectral radius restriction creates runtimes $O(100)$ larger
  than the production Krylov methods for the fuel assembly criticality
  problem
\end{itemize}
